---
title: "Song Sparrow Dataset Analysis"
author: "TO MINH ANH"
format: pdf
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import libraries and load the dataset

```{r}
# Load necessary libraries
library(ggplot2)
library(patchwork)  # For side-by-side plots
library(gridExtra)
library(dplyr)
library(lme4)
library(nlme)

# Install and load the car package
#install.packages("car")
library(car)
```


```{r}
# Load the dataset
data = read.csv("female.csv")

# Summary/Structure of the dataset
summary(data)
#glimpse(data)
```

```{r}
length(unique(data$band))
```

The data set comprises 742 entries and 8 columns, with data corresponding to 360 unique sparrows included in the study. In this analysis, `year` and `cohort` are converted into categorical variables.


```{r}
# Convert year and cohort to factors
data$year = as.factor(data$year)
data$cohort = as.factor(data$cohort)
```

# Study Design

```{r}
# Handle missing values
data_clean = na.omit(data)
```


# Data Description


## Visualizing the Relationship Between the Number of Offspring and Grouping Factors

```{r}
# Ensure 'year' and 'cohort' are numeric (convert factor to numeric if necessary)
data$year = as.numeric(as.character(data$year))
data$cohort = as.numeric(as.character(data$cohort))

# Drop missing values for 'year' and 'cohort'
data_clean = data[!is.na(data$year) & !is.na(data$cohort), ]

# Count sample sizes for each year
year_counts = as.data.frame(table(data_clean$year))
colnames(year_counts) = c("year", "sample_size")
year_counts$year = as.numeric(as.character(year_counts$year))

# Count sample sizes for each cohort
cohort_counts = as.data.frame(table(data_clean$cohort))
colnames(cohort_counts) = c("cohort", "sample_size")
cohort_counts$cohort = as.numeric(as.character(cohort_counts$cohort))

# Create the plot for year
plot_year = ggplot(year_counts, aes(x = as.factor(year), y = sample_size)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Sample Sizes Across Year", x = "Year", y = "Sample Size") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Create the plot for cohort
plot_cohort = ggplot(cohort_counts, aes(x = as.factor(cohort), y = sample_size)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Sample Sizes Across Cohort", x = "Cohort", y = "Sample Size") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Combine the two plots side by side
combined_plot = plot_year + plot_cohort

# Print the combined plot
print(combined_plot)
```

```{r}
# Convert year and cohort to factors
data$year = as.factor(data$year)
data$cohort = as.factor(data$cohort)
```

```{r}
# Box plot of spf by Year (Macro Grouping Factor)
ggplot(data, aes(x = as.factor(year), y = spf)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(
    title = "Number of Offspring (spf) by Year",
    x = "Year",
    y = "Number of Offspring (spf)"
  ) +
  theme_minimal()

# Box plot of spf by Cohort (Can be Macro or Micro Grouping Factor)
ggplot(data, aes(x = as.factor(cohort), y = spf)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  labs(
    title = "Number of Offspring (spf) by Cohort",
    x = "Cohort",
    y = "Number of Offspring (spf)"
  ) +
  theme_minimal()

# Box plot of spf by Bird ID (Micro Grouping Factor)
# Since there may be too many birds, we'll limit to a subset of birds
subset_birds = data %>%
  filter(band %in% sample(unique(band), 20))  # Sample 20 random birds for visualization

ggplot(subset_birds, aes(x = as.factor(band), y = spf)) +
  geom_boxplot(fill = "pink", color = "darkred") +
  labs(
    title = "Number of Offspring (spf) by Bird ID (Subset of 20 Birds)",
    x = "Bird ID (Band)",
    y = "Number of Offspring (spf)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```

## Relationship Between Cohort, Year, and Age

Check if all the values in the columns `age`, `year`, and `cohort` follow the formula `age=year−cohort+1`

```{r}
# Ensure year, cohort, and age are numeric (in case they are read as factors)
data$year = as.numeric(as.character(data$year))
data$cohort = as.numeric(as.character(data$cohort))
data$age = as.numeric(as.character(data$age))

# Remove rows with missing values in age, year, or cohort
filtered_data = na.omit(data[, c("age", "year", "cohort")])

# Check if the formula holds for all rows
valid_formula = filtered_data$age == (filtered_data$year - filtered_data$cohort + 1)

# Check if all the values satisfy the condition
all_valid = all(valid_formula, na.rm = TRUE)

# Print result
if (all_valid) {
  print("All non-missing rows satisfy the formula age = year - cohort + 1")
} else {
  print("There are rows where the formula does not hold.")
}

# See which rows violate the formula, print them
violating_rows = filtered_data[!valid_formula, ]
if (nrow(violating_rows) > 0) {
  print("Rows that do not satisfy the formula:")
  print(violating_rows)
}
```

```{r}
# Get all rows with at least one missing value
rows_with_missing = data[!complete.cases(data), ]

# Count the number of rows with missing values
num_rows_with_missing = nrow(rows_with_missing)
print(paste("Number of rows with at least one missing value:", num_rows_with_missing))
```

```{r}
# Create a table
table_data = table(data$year, data$cohort)
print(table_data)
# Calculate the correlation between year and cohort
cor(as.numeric(data$year), as.numeric(data$cohort), use = "complete.obs")
cor(as.numeric(data$year), as.numeric(data$age), use = "complete.obs")
cor(as.numeric(data$cohort), as.numeric(data$age), use = "complete.obs")
```

```{r}
# Convert table to data frame
df_table = as.data.frame(table(data$year, data$cohort))

# Plot as heatmap
ggplot(df_table, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Heatmap of Year vs Cohort",
       x = "Cohort",
       y = "Year",
       fill = "Frequency") +
  theme_minimal()
```


## Preliminary Modeling


### Sequential Hypothesis Testing

**Micro Variables**

```{r}
# Fit the full model
lm_full = lm(spf ~ age + fpop + x + y, data = data)
# Set contrasts to be compatible with Type III ANOVA
options(contrasts = c("contr.sum", "contr.poly"))

# Perform Type III ANOVA
Anova(lm_full, type = "III")
```
```{r}
summary(lm_full)
```

```{r}
vif(lm_full)  # Variance Inflation Factor
```

**Macro Variables**

```{r}
lm = lm(spf ~ fpop + x + age, data = data_clean)

lm_inter1 = lm(spf ~ fpop + x + age + fpop:age, data = data_clean)
lm_inter2 = lm(spf ~ fpop + x + age + x:y, data = data_clean)

lm_cohort = lm(spf ~ fpop + x + age + as.factor(cohort), data = data_clean)
lm_year = lm(spf ~ fpop + x + age + as.factor(year), data = data_clean)
lm_both = lm(spf ~ fpop + x + age + as.factor(year) + as.factor(cohort), data = data_clean)

anova(lm, lm_year)
anova(lm, lm_cohort)
anova(lm_cohort, lm_both)  
anova(lm_year, lm_both)   

AIC(lm, lm_inter1, lm_inter2, lm_cohort, lm_year, lm_both)
BIC(lm, lm_inter1, lm_inter2, lm_cohort, lm_year, lm_both)
```

### Checking assumptions

```{r}
# Diagnostic plots for assumptions
par(mfrow = c(1, 2))  # Arrange plots in a grid
plot(lm_year)  # Residuals vs fitted, Q-Q plot, Scale-location, Residuals leverage
```
    
# Model fitting and diagnostics

## Linear Mixed-Effects Models

```{r}
fit1 = lmer(spf ~ fpop + x + age + (1 | band) + (1 | year), data = data_clean)
fit2 = lmer(spf ~ fpop + x + age + (1 | year), data = data)
fit3 = lmer(spf ~ fpop + x + age + (1 | band), data = data)
BIC(fit1, fit2, fit3)
```

## Modeling Temporal Correlation

```{r}
# Handle missing values
data_clean = na.omit(data)
```

```{r}
# Fit the corAR1 model with year as the grouping factor
fit4 = lme(
  spf ~ fpop + x + age,                # Fixed effects
  random = ~1 | year,                  # Random intercept for year
  correlation = corAR1(),              # AR(1) structure for temporal correlation
  data = data_clean,                   # Cleaned dataset
  method = "ML"                        # Use Maximum Likelihood for comparison
)

# Check BIC of the model
BIC(fit4)
```

```{r}
# Fit a model with random slopes for age by year
fit5 = lme(
  spf ~ fpop + x + age,                 # Fixed effects
  random = list(year = pdDiag(~ age)),  # Random intercept and slope for age by year
  correlation = corAR1(),               # AR(1) structure for temporal correlation
  data = data_clean,
  method = "ML"
)

# Compare BIC with fit_corAR1
BIC(fit4, fit5)
```

Checking interactions

```{r}
add1(fit4, "fpop*age", data = data, test = "Chisq")
add1(fit4, "fpop*x", data = data, test = "Chisq")
add1(fit4, "x*y", data = data, test = "Chisq")
add1(fit4, "fpop*y", data = data, test = "Chisq")
add1(fit4, "age*y", data = data, test = "Chisq")
add1(fit4, "age*x", data = data, test = "Chisq")
add1(fit4, "band*age", data = data, test = "Chisq")
add1(fit4, "band*x", data = data, test = "Chisq")
add1(fit4, "band*fpop", data = data, test = "Chisq")
```

## Checking assumptions

```{r}
par(mfrow = c(1, 2)) 

# Generate plots
qqnorm(resid(fit4), main = "Normal Q-Q Plot")
qqline(resid(fit4), col = "red")

hist(resid(fit4), breaks = 10, main = "Histogram of Residuals", xlab = "Residuals")
```

```{r}
# Set up a 2x2 plotting canvas
par(mfrow = c(1, 2))

acf(resid(fit4), main = "ACF of Residuals")

plot(fitted(fit4), residuals(fit4, type = "pearson"), 
     main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")
```

# Model and data analysis interpretation

```{r}
# Summary of the final model
summary(fit4)
```

```{r}
# Compute confidence intervals for the model
intervals(fit4, level = 0.95)  # Default is 95% confidence level
```

```{r}
# Install from CRAN
install.packages("glmmTMB")

# If you need lme4 as well
install.packages("lme4")

# Optional but recommended: install for model comparison and diagnostics
install.packages(c("performance", "DHARMa"))

# Load the packages after installation
library(glmmTMB)
library(lme4)
library(performance)  # for model diagnostics
library(DHARMa)       # for residual diagnostics
```

```{r}
# Poisson mixed-effects model
fit_poisson = glmmTMB(
  spf ~ fpop + x + age +  # Fixed effects
    (1 | year),           # Random intercept for year
  family = poisson(),     # Poisson distribution
  data = data_clean,
  REML = FALSE            # Equivalent to ML method
)
```



